---
title: "Hierarchical Logistic Growth Curves"
author: "Brynjólfur Gauti Jónsson"
date: "3/26/2020"
output: 
    html_document:
        theme: flatly
        toc: true
        toc_float: true
        code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = F, message = F, 
                      fig.asp = 0.621, out.width = "100%", fig.width = 8)
```

```{r}
library(tidyverse); library(knitr); library(kableExtra); library(broom); library(cowplot); 
library(rstan); library(tidybayes); library(scales); library(DT); library(plotly)



theme_set(theme_classic(base_size = 12) + 
            background_grid(color.major = "grey90", 
                            color.minor = "grey95", 
                            minor = "xy", major = "xy") +
            theme(legend.position = "none"))
m <- read_rds("Hierarchical_Model_NegBin.rds")
d <- read_csv("Interactive Model Checking/stan_data.csv")
countries <- d %>% distinct(country, country_id)
which_iceland <- d %>% filter(country == "Iceland") %>% .$country_id %>% unique
n_countries <- max(d$country_id)
```

```{r, include = F, eval = F}
# library(shinystan)
# stan_obj <- launch_shinystan(m)
# deploy_shinystan(stan_obj, appName = "LogisticGrowthCurves", account = "bgautijonsson")
```


```{r}
results <- tidyMCMC(m, conf.int = T, rhat = T, ess = T, estimate.method = "median", conf.method = "quantile") %>% 
  mutate(par = str_match(term, "[a-z_2]+")) %>% 
  group_by(par) %>% 
  mutate(num = row_number() %>% as.numeric)
```

# Updates

## 2020-03-26

* Implemented a negative binomial likelihood with country-specific overdispersion.
* Changed filtering criteria for inclusion in model data in order to include more countries.
* Did some visual inspection of data in order to make sure we don't lose countries to inner_join strategies. Saw that we were missing USA among others due to different names in different datasets, but is now fixed.

# Methods

## Parametrization

Let $E_i$ and $I_{i, t}$ be the population and number of infected in country $i$ at time $t$. Then the percent of infected can be calculated as

$$
P_{i, t} = \frac{I_{i, t}}{E_i}.
$$

In the usual Logistic Regression GLM we could model the percent of infected, as a function of time *(in days)*, with

$$
\log\left(\frac{P_{i, t}}{1 - P_{i, t}}\right) = \alpha_i + \beta_i \cdot t,
$$

where $\alpha_i$ is a measure of how many have been infected in country $i$ at time $t = 0$ and $\beta_i$ is a measure of growth. In the case of COVID-19 infections we don't know the maximum percent of populations that will be infected, so we have another unknown parameter, the saturation percent at which a country will reach its maximum number of infected, $S_i$. Thus our model looks like

$$
\log\left(\frac{P_{i, t}}{S_i - P_{i, t}}\right) = \alpha_i + \beta_i \cdot t.
$$

These parameters are hard to estimate when data from only one country are used. However, if we were to pool information about them between countries, as in a hierarchical Bayesian model, estimation might be possible. Let

$$
z_{i, t} = \alpha_i + \beta_i \cdot t,
$$

where $\alpha_i$ is a measure of how many have been infected in country $i$ at time $t = 0$ and $\beta_i$ is a measure of growth. Then

$$
P_{i, t} = \frac{S_i}{1 + \exp(-z_{i, t})},
$$

and conditional on some sampling distribution, $f$, we could write

$$
I_{i, t} \sim \mathrm{f_\theta}(P_{i, t},  E_i),
$$

where $\theta$ contains all relevant parameters to be estimated.

## Bayesian Inference

Bayesian inference is a great tool when small amounts of data are to be shared from various sources. In this case the sources are different countries, and the data are cumulative numbers of cases. If we utilize a Negative Binomial likelihood for the observed cumulative cases, then

$$
I_{i, t} \sim \mathrm{NegBin}(P_{i, t} \cdot E_i, \phi_i),
$$

where $\phi_i$ is a country-level effect specifying the amount of overdispersion in that country.

### Modeling daily counts

However, there is a lot of implicit correlation in the values of $I_{i, t}$ for different values of $t$. Thus, a better parametrization would be to model the daily number of cases. 

Let

$$
z_{i, t} = \alpha_i + \beta_i \cdot t,
$$

so that the percent of infected, $P_i$, is

$$
P_{i, t} = \frac{S_i}{1 + \exp(-z_{i,t})}.
$$

If we furthermore write

$$
z^*_{i, t - 1} = \alpha_i + \beta_i \cdot (t - 1),
$$

and

$$
P^*_{i, t - 1} = \frac{S_i}{1 + \exp(-z^*_{i, t-1})},
$$

the change in rates between days is

$$
C_{i, t} = P_{i, t} - P^*_{i, t - 1}.
$$

Since $C_{i, t}$ is simply the first derivative of $P_{i, t}$ with respect to $t$, we can skip the differencing step and directly model the derivative

$$
C_{i, t} = \frac{d}{dt}P_{i, t} = \beta_i S_i \frac{\exp{(-z_{i, t})}}{(\exp(-z_{i, t}) + 1))^2}
$$

Then, conditional on a Negative Binomial likelihood and population size $E_i$, the daily number of observed cases, $D_{i, t}$, can be written as

$$
D_{i, t} \sim \mathrm{NegBin}(C_{i, t} \cdot E_i, \phi_i)
$$

### Parameters

The parameters, $\alpha$ and $\beta$, are treated as in a generalized linear model, except that we use the constraints $\beta_i > 0$ and $\mu_\beta > 0$. We put hierarchical priors on $\alpha_i$ and $\beta_i$ so that for each country, $i$,

$$
\begin{aligned}
\beta_i &\sim \mathrm{Normal}(\mu_\beta, \sigma^2_\beta) \\
\alpha_i &\sim \mathrm{Normal}(\mu_\alpha, \sigma^2_\alpha)
\end{aligned}
$$

The $\mu$ parameters are given a flat prior distribution, while $\sigma^2_\beta$ and $\sigma^2_\alpha$ are given $\mathrm{Inv-}\chi^2(2)$ prior distributions. This is equal to putting $\mathrm{Exponential}(\frac12)$ prior distributions on the precisions $\frac{1}{\sigma^2_\beta}$ and $\frac{1}{\sigma^2_\alpha}$. 

The $S_i$ parameters take on values in $(0, 1)$, so we thought it proper to model them with Beta distributions. By putting hierarchical priors on them we could also share information between countries on the estimated saturation points. Thus the saturation parameter for country $i$ is sampled as

$$
S_i \sim \mathrm{Beta}(a_S, b_S),
$$

where $a_S$ and $b_S$, being strictly positive, are given flat prior distributions on the $\log$ scale, which is equal to $p(a_S) \propto \frac{1}{a_S}$ and $p(b_S) \propto \frac{1}{b_S}$.

We parametrize the negative binomial likelihood in the form of mean and overdispersion. If we write 

$$
D_{i, t} \sim \mathrm{NegBin}(\mu_{i, t}, \phi_i),
$$

where

$$
\mu_{i, t} = C_{i, t} \cdot E_i,
$$

we can write

$$
E[D_{i, t}] = \mu_{i, t} \qquad Var[D_{i, t}] = \mu_{i, t} + \phi_i \cdot \mu_{i, t}^2.
$$

Using this parametrization we put a hierarchical exponential prior on the $\phi_i$ parameters,

$$
\phi_i \sim \mathrm{Exponential}(\lambda_\phi).
$$
Since $\lambda_\phi$ is strictly positive we give it a flat prior on the log scale, or $p(\lambda_\phi) \propto \frac{1}{\lambda_\phi}$

***

**Putting it all together we get**

$$
\begin{aligned}
D_{i, t} &\sim \mathrm{NegBin}(C_{i, t} \cdot E_i, \phi_i) \\
C_{i, t} = \frac{d}{dt}P_{i, t} &= \beta_i S_i \frac{\exp{(-z_{i, t})}}{(\exp(-z_{i, t}) + 1))^2} \\
z_{i, t} &= \alpha_i + \beta_i \cdot t \\
\beta_i &\sim \mathrm{Normal}(\mu_\beta, \sigma^2_\beta) \\
\alpha_i &\sim \mathrm{Normal}(\mu_\alpha, \sigma^2_\alpha) \\
S_i &\sim \mathrm{Beta}(a_S, b_S) \\
\phi_i &\sim \mathrm{Exponential}(\lambda_\phi)\\
p(\lambda_\phi) &\propto \frac{1}{\lambda_\phi} \\
p(\mu_\alpha) = p(\mu_\beta)  &\propto 1 \\
p(\frac{1}{\sigma^2_\alpha}) = p(\frac{1}{\sigma^2_\beta}) &\sim \mathrm{Exponential}(\frac12)  \\
p(a_S) &\propto \frac{1}{a_S} \\
p(b_S) &\propto \frac{1}{b_S}
\end{aligned}
$$

***

## Data

Data used for modeling were filtered according to

* Only keep data for which cases per 1000 inhabitants $\geq$ 0.02.
* Only keep data for countries for which the first observation after this filtering is $\leq$ 0.06
* Only keep data for countries with more than 4 days of follow-up after the above steps.

```{r}
d %>% 
  group_by(country) %>% 
  summarise(First = min(date),
            Days_In_Data = n(),
            Start_Rate = min(case_rate),
            End_Rate = max(case_rate)) %>% 
  set_names(c("Country", "Entry", "Days in data", "Start", "End")) %>% 
  kable(caption = "Table 1. Summary information about countries used in modeling",
        align = c("l", rep("c", ncol(.) - 1)),
        digits = 3) %>% 
  kable_styling(bootstrap_options = c("striped", "hover")) %>% 
  row_spec(which_iceland, bold = T) %>% 
  add_header_above(c("", "", "", "Rate per 1000" = 2)) %>%
  scroll_box(height = "500px")
```


```{r, fig.asp = 0.8}
p <- d %>% 
  ggplot(aes(days, case_rate, group = country, col = country == "Iceland")) +
  geom_line() +
  scale_y_log10() +
  scale_colour_manual(values = c("grey", "blue")) +
  labs(x = "Days since rate reached 0.02 per 1000",
       y = "Cases per 1000",
       title = "Observed trends for countries used in data",
       subtitle = "Shown as days since a country entered the modeling data")
ggplotly(p)
```

## Software

The model is fit using [Stan's](https://mc-stan.org/) R interface.

All code is available at https://github.com/bgautijonsson/covid19. We're sorry for the mess and we're working on it! We are also going to translate the README files into English.

# Results

## Convergence

### Sampler configuration

Eight chains were run for 3000 iterations after a warm-up period of 3000 iterations. The slowest chain converged in 280 seconds. 

### Warnings

* There were 67 divergent transitions after warmup.
* Bulk effective sample size was too low.
* Tail effective sample size was too low.




## Country Level Effects

```{r}
results %>% 
  ungroup %>% 
  filter(par %in% c("beta", "beta2", "alpha", "maximum", "phi_inv", "detected_perc")) %>%
  mutate(par = str_replace(par, "phi_inv", "phi")) %>% 
  inner_join(countries, by = c("num" = "country_id")) %>% 
  mutate(par = str_to_title(par)) %>% 
  select(par, country, num, everything(), -num, -term, -std.error) %>% 
  set_names(c("Parameter", "Country", "Median", "Lower", "Upper", "Rhat", "ESS")) %>% 
  kable(digits = 4, align = c("l", "l", rep("c", ncol(.) - 2)),
        caption = "Table 2. Summary of posterior samples of country level effects") %>% 
  kable_styling(bootstrap_options = c("striped", "hover")) %>% 
  add_header_above(c("", "", "", "95% PI" = 2, "Convergence" = 2)) %>% 
  column_spec(1, bold = T) %>% 
  row_spec(which_iceland + c(0, 1, 2) * n_countries, bold = T) %>% 
  collapse_rows(1, valign = "top") %>% 
  scroll_box(height = "600px")
```

```{r, fig.width = 9, fig.asp = 1.5}
p <- results %>% 
  ungroup %>% 
  filter(par %in% c("beta", "alpha", "beta2", "maximum", "phi_inv", "detected_perc", "beta2")) %>% 
  mutate(par = str_replace(par, "phi_inv", "phi")) %>% 
  inner_join(countries, by = c("num" = "country_id")) %>% 
  mutate(par = str_to_title(par)) %>% 
  mutate(plot_var = str_c(par, "_", country)) %>% 
  mutate(country = fct_reorder(country, estimate)) %>% 
  ggplot(aes(country, estimate, ymin = conf.low, ymax = conf.high, col = country == "Iceland")) +
  geom_linerange() +
  geom_point() +
  coord_flip() +
  facet_wrap("par", scales = "free_x") +
  scale_colour_manual(values = c("grey", "blue")) +
  labs(title = "Posterior medians and 95% PIs") +
  theme(axis.title = element_blank())

ggplotly(p)
```

## Hyperparameters

```{r}
results %>% 
  ungroup %>% 
  filter(par %in% c("mu_beta", "sigma_sq_beta", "sigma_sq_beta2", 
                    "mu_alpha", "sigma_sq_alpha", "beta_a", "beta_b",
                    "mu_phi_inv", "detect_a", "detect_b",
                    "mu_beta2", "sigma_sq_beta2", "mu_error", "sigma_sq_error")) %>% 
  mutate(par = str_replace(par, "mu_phi_inv", "mu_phi"),
         par = str_to_title(par)) %>% 
  select(par, everything(), -num, -term, -std.error) %>% 
  set_names(c("Parameter",  "Median", "Lower", "Upper", "Rhat", "ESS")) %>% 
  kable(digits = 4, align = c("l", "l", rep("c", ncol(.) - 2)),
        caption = "Table 3. Summary of posterior samples of hyperparameters") %>% 
  kable_styling(bootstrap_options = c("striped", "hover")) %>% 
  add_header_above(c("", "", "95% PI" = 2, "Convergence" = 2)) %>% 
  column_spec(1, bold = T)
```

## Posterior Predictions

We obtain predictions by conditioning on the first observed number of cumulative cases in a country (as seen in the modeling data), perform simulations of new daily cases and sum those up to get predicted cumulative cases.

```{r}
knitr::include_app("https://bgautijonsson.shinyapps.io/Hierarchical_Report_NegBin/", height = "600px")
```

# Some things to keep in mind

## Comparisons between countries

This model does not predict the total amount of COVID-19 cases, it predicts the total amount of DIAGNOSED COVID-19 cases. It then depends on each country's specific diagnostic criteria what that number means. Iceland does many tests per capita so there it is predicting a broader spectrum of cases, not all serious, whereas in a country like Italy or Spain the model's predictions have another meaning as the diagnostic criteria are different from those of Iceland.

## Changes in policies

As of yet, the model only implements one slope for each country, which is assumed to be fixed throughout the epidemic. Each country's policies should of course affect this growth rate and the date at which a country implements a policy should affect when the slope changes. Finding solutions to this is on our next actions list.

# Next actions

## Definite

* Time-varying growth rates.
* Include country-level information in the model

## Maybe

* Include information on number of recovered to estimate country-specific recovery rates conditional on being diagnosed.
* Include information on deaths to estimate country-specific death rates conditional on being diagnosed.

# Appendix

## Stan Code

```{r, eval = FALSE}
data {
  int<lower = 0> N_obs;
  int country[N_obs];
  vector[N_obs] days;
  int new_cases[N_obs];
  int<lower = 0> N_countries;
  vector[N_countries] pop;
}

parameters {
  vector<lower = 0>[N_countries] beta;
  vector[N_countries] alpha;
  vector<lower = 0, upper = 1>[N_countries] maximum;
  
  real<lower = 0> mu_beta;
  real<lower = 0> sigma_sq_beta;
  
  real mu_alpha;
  real<lower = 0> sigma_sq_alpha;
  
  real<lower = 0> beta_a;
  real<lower = 0> beta_b;
  
  vector<lower = 0>[N_countries] phi_inv;
  real<lower = 0> mu_phi_inv;
  
}

transformed parameters {
  vector[N_obs] linear = alpha[country] + beta[country] .* days;
  vector<lower = 0, upper = 1>[N_obs] difference;
  vector<lower = 0>[N_countries] phi = inv(phi_inv);
  for (i in 1:N_obs) {
    difference[i] = beta[country[i]] * maximum[country[i]] * exp(-linear[i]) / square(exp(-linear[i]) + 1);
  }
}

model {
  
  sigma_sq_beta ~ inv_chi_square(2);
  sigma_sq_alpha ~ inv_chi_square(2);
  
  maximum ~ beta(beta_a, beta_b);
  
  beta ~ normal(mu_beta, sigma_sq_beta);
  alpha ~ normal(mu_alpha, sigma_sq_alpha);
  
  phi_inv ~ exponential(mu_phi_inv);
  
  new_cases ~ neg_binomial_2(difference .* pop[country], phi[country]);
}


```

