---
title: "Hierarchical Logistic Growth Curves"
author: "Brynjólfur Gauti Jónsson"
date: "3/24/2020"
output: 
    html_document:
        theme: flatly
        toc: true
        toc_float: true
        code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = F, message = F, 
                      fig.asp = 0.621, out.width = "100%", fig.width = 8)
```

```{r}
library(tidyverse); library(knitr); library(kableExtra); library(broom); library(cowplot); 
library(rstan); library(tidybayes); library(scales); library(DT); library(plotly)

theme_set(theme_classic(base_size = 12) + 
            background_grid(color.major = "grey90", 
                            color.minor = "grey95", 
                            minor = "xy", major = "xy") +
            theme(legend.position = "none"))
m <- read_rds("Hierarchical_Model_NegBin.rds")
d <- read_csv("Interactive Model Checking/stan_data.csv")
countries <- d %>% distinct(country, country_id)
which_iceland <- d %>% filter(country == "Iceland") %>% .$country_id %>% unique
n_countries <- max(d$country_id)
```

```{r}
results <- tidyMCMC(m, conf.int = T, rhat = T, ess = T, estimate.method = "median", conf.method = "quantile") %>% 
  mutate(par = str_match(term, "[a-z_2]+")) %>% 
  group_by(par) %>% 
  mutate(num = row_number() %>% as.numeric)
```

# Methods

## Parametrization

Let $E_i$ and $I_{i, t}$ be the population and number of infected in country $i$ at time $t$. Then the percent of infected can be calculated as

$$
P_{i, t} = \frac{I_{i, t}}{E_i}.
$$

Let

$$
z_{i, t} = \alpha_i + \beta_i \cdot t,
$$

where $\alpha_i$ is a measure of how many have been infected in country $i$ at time $t = 0$ and $\beta_i$ is a measure of growth. Then

$$
P_{i, t} = \frac{S_i}{1 + \exp(-z_{i, t})},
$$

and conditional on some sampling distribution, $f$, we could write

$$
I_{i, t} \sim \mathrm{f_\theta}(P_{i, t},  E_i),
$$

where $\theta$ contains all parameters to be estimated.

## Bayesian Inference

Bayesian inference is a great tool when small amounts of data are to be shared from various sources. In this case the sources are different countries, and the data are cumulative numbers of cases. If we utilize a Negative Binomial likelihood for the observed cumulative cases, then

$$
I_{i, t} \sim \mathrm{NegBin}(P_{i, t} \cdot E_i, \phi_i),
$$

where $\phi_i$ is a country-level effect specifying the amount of overdispersion in that country.

### Modeling daily counts

However, there is a lot of implicit correlation in the values of $I_{i, t}$ for different values of $t$. Thus, a better parametrization would be to model the daily number of cases. 

Let

$$
z_{i, t} = \alpha_i + \beta_i \cdot t,
$$

so that the percent of infected, $P_i$, is

$$
P_{i, t} = \frac{S_i}{1 + \exp(-z_{i,t})}.
$$

If we furthermore write

$$
z^*_{i, t - 1} = \alpha_i + \beta_i \cdot (t - 1),
$$

and

$$
P^*_{i, t - 1} = \frac{S_i}{1 + \exp(-z^*_{i, t-1})},
$$

the change in rates between days is

$$
C_{i, t} = P_{i, t} - P^*_{i, t - 1}.
$$

Since $C_{i, t}$ is simply the first derivative of $P_{i, t}$ with respect to $t$, we can skip the differencing step and directly model the derivative

$$
C_{i, t} = \frac{d}{dt}P_{i, t} = \beta_i S_i \frac{\exp{(-z_{i, t})}}{(\exp(-z_{i, t}) + 1))^2}
$$

Then, conditional on a Negative Binomial likelihood and population size $E_i$, the daily number of observed cases, $D_{i, t}$, can be written as

$$
D_{i, t} \sim \mathrm{NegBin}(C_{i, t} \cdot E_i, \phi_i)
$$

### Parameters

The parameters, $\alpha$ and $\beta$, are treated as in a generalized linear model, and we put a hierarchical prior on them so that for each country, $i$,

$$
\begin{aligned}
\beta_i &\sim \mathrm{Normal}(\mu_\beta, \sigma^2_\beta) \\
\alpha_i &\sim \mathrm{Normal}(\mu_\alpha, \sigma^2_\alpha)
\end{aligned}
$$

The $\mu$ parameters are given informative prior distributions based on our previously fitted *Poisson* model. They are as follows 

$\sigma^2_\beta$ and $\sigma^2_\alpha$ are given $\mathrm{Inv-}\chi^2(2)$ prior distributions. This is equal to putting $\mathrm{Exponential}(\frac12)$ prior distributions on the precisions $\frac{1}{\sigma^2_\beta}$ and $\frac{1}{\sigma^2_\alpha}$. 

The $S_i$ parameters take on values in $(0, 1)$, so we thought it proper to model them with Beta distributions. By putting hierarchical priors on them we could also share information between countries on the estimated saturation points. Thus the saturation parameter for country $i$ is sampled as

$$
S_i \sim \mathrm{Beta}(a_S, b_S),
$$

where $a_S$ and $b_S$, being strictly positive, are given flat prior distributions on the $\log$ scale, which is equal to $p(a_S) \propto \frac{1}{a_S}$ and $p(b_S) \propto \frac{1}{b_S}$.

We parametrize the negative binomial likelihood in the form of mean and overdispersion. If we write 

$$
D_{i, t} \sim \mathrm{NegBin}(\mu_{i, t}, \phi_i),
$$

where

$$
\mu_{i, t} = C_{i, t} \cdot E_i,
$$

we can write

$$
E[D_{i, t}] = \mu_{i, t} \qquad Var[D_{i, t}] = \mu_{i, t} + \phi_i \cdot \mu_{i, t}^2.
$$

Using this paremetrization we put hierarchical exponential priors on the $\phi_i$ parameters, or

$$
\phi_i \sim \mathrm{Exponential}(\lambda_\phi) \\
\lambda_\phi \sim \mathrm{Exponential}(1)
$$

***

**Putting it all together we get**

$$
\begin{aligned}
D_{i, t} &\sim \mathrm{NegBin}(C_{i, t} \cdot E_i, \phi_i) \\
C_{i, t} = \frac{d}{dt}P_{i, t} &= \beta_i S_i \frac{\exp{(-z_{i, t})}}{(\exp(-z_{i, t}) + 1))^2} \\
z_{i, t} &= \alpha_i + \beta_i \cdot t \\
\beta_i &\sim \mathrm{Normal}(\mu_\beta, \sigma^2_\beta) \\
\alpha_i &\sim \mathrm{Normal}(\mu_\alpha, \sigma^2_\alpha) \\
S_i &\sim \mathrm{Beta}(a_S, b_S) \\
p(\mu_\alpha) = p(\mu_\beta)  &\propto 1 \\
p(\frac{1}{\sigma^2_\alpha}) = p(\frac{1}{\sigma^2_\beta}) &\sim \mathrm{Exponential}(\frac12)  \\
p(a_S) &\propto \frac{1}{a_S} \\
p(b_S) &\propto \frac{1}{b_S}
\end{aligned}
$$

***

## Data

Data used for modeling were filtered according to

* Only keep data for which cases per 1000 inhabitants $\geq$ 0.02.
* Only keep data for countries for which the first observation after this filtering is $\leq$ 0.06
- This reduces instability in estimates. Some very small countries, such as San Marino, tend to have large jumps in case rates. Other reasons for large jumps are governments starting large scale testing.
* Only keep data for countries with more than 8 days of follow-up after the above steps.

```{r}
d %>% 
  group_by(country) %>% 
  summarise(First = min(date),
            Days_In_Data = n(),
            Start_Rate = min(case_rate),
            End_Rate = max(case_rate)) %>% 
  set_names(c("Country", "Entry", "Days in data", "Start", "End")) %>% 
  kable(caption = "Table 1. Summary information about countries used in modeling",
        align = c("l", rep("c", ncol(.) - 1)),
        digits = 3) %>% 
  kable_styling(bootstrap_options = c("striped", "hover")) %>% 
  row_spec(which_iceland, bold = T) %>% 
  add_header_above(c("", "", "", "Rate per 1000" = 2)) %>%
  scroll_box(height = "500px")
```


```{r, fig.asp = 0.8}
p <- d %>% 
  ggplot(aes(days, case_rate, group = country, col = country == "Iceland")) +
  geom_line() +
  scale_y_log10() +
  scale_colour_manual(values = c("grey", "blue")) +
  labs(x = "Days since rate reached 0.02 per 1000",
       y = "Cases per 1000",
       title = "Observed trends for countries used in data",
       subtitle = "Shown as days since a country entered the modeling data")
ggplotly(p)
```


# Results


## Country Level Effects

```{r}
results %>% 
  ungroup %>% 
  filter(par %in% c("beta", "beta2", "alpha", "maximum", "phi_inv", "detected_perc")) %>%
  mutate(par = str_replace(par, "phi_inv", "phi")) %>% 
  inner_join(countries, by = c("num" = "country_id")) %>% 
  mutate(par = str_to_title(par)) %>% 
  select(par, country, num, everything(), -num, -term, -std.error) %>% 
  set_names(c("Parameter", "Country", "Median", "Lower", "Upper", "Rhat", "ESS")) %>% 
  kable(digits = 4, align = c("l", "l", rep("c", ncol(.) - 2)),
        caption = "Table 2. Summary of posterior samples of country level effects") %>% 
  kable_styling(bootstrap_options = c("striped", "hover")) %>% 
  add_header_above(c("", "", "", "95% PI" = 2, "Convergence" = 2)) %>% 
  column_spec(1, bold = T) %>% 
  row_spec(which_iceland + c(0, 1, 2) * n_countries, bold = T) %>% 
  collapse_rows(1, valign = "top") %>% 
  scroll_box(height = "600px")
```

```{r, fig.width = 9, fig.asp = 1.5}
p <- results %>% 
  ungroup %>% 
  filter(par %in% c("beta", "alpha", "beta2", "maximum", "phi_inv", "detected_perc", "beta2")) %>% 
  mutate(par = str_replace(par, "phi_inv", "phi")) %>% 
  inner_join(countries, by = c("num" = "country_id")) %>% 
  mutate(par = str_to_title(par)) %>% 
  mutate(plot_var = str_c(par, "_", country)) %>% 
  mutate(country = fct_reorder(country, estimate)) %>% 
  ggplot(aes(country, estimate, ymin = conf.low, ymax = conf.high, col = country == "Iceland")) +
  geom_linerange() +
  geom_point() +
  coord_flip() +
  facet_wrap("par", scales = "free_x") +
  scale_colour_manual(values = c("grey", "blue")) +
  labs(y = "Posterior median and 95% PI") +
  theme(axis.title.y = element_blank())

ggplotly(p)
```

## Hyperparameters

Our method of filtering the data used in modeling ensured that countries have similar starting percentages of infected, so the $\alpha$ parameters can be more readily pooled. However, some smalelr countries that are used tend to have large jumps in rates dues to their small populations. Thus the variance in the alpha parameters can be rather large. Our way of filtering data could still be improved to increase stability in the $\alpha$ parameters.

```{r}
results %>% 
  ungroup %>% 
  filter(par %in% c("mu_beta", "sigma_sq_beta", "sigma_sq_beta2", 
                    "mu_alpha", "sigma_sq_alpha", "beta_a", "beta_b",
                    "mu_phi_inv", "detect_a", "detect_b",
                    "mu_beta2", "sigma_sq_beta2", "mu_error", "sigma_sq_error")) %>% 
  mutate(par = str_replace(par, "mu_phi_inv", "mu_phi"),
         par = str_to_title(par)) %>% 
  select(par, everything(), -num, -term, -std.error) %>% 
  set_names(c("Parameter",  "Median", "Lower", "Upper", "Rhat", "ESS")) %>% 
  kable(digits = 4, align = c("l", "l", rep("c", ncol(.) - 2)),
        caption = "Table 3. Summary of posterior samples of hyperparameters") %>% 
  kable_styling(bootstrap_options = c("striped", "hover")) %>% 
  add_header_above(c("", "", "95% PI" = 2, "Convergence" = 2)) %>% 
  column_spec(1, bold = T)
```

## Posterior Predictions

We obtain predictions by conditioning on the first observed number of cumulative cases in a country (as seen in the modeling data), perform simulations of new daily cases and sum those up to get predicted cumulative cases.

```{r}
knitr::include_app("https://bgautijonsson.shinyapps.io/Hierarchical_Report/", height = "600px")
```


# Next actions


## Time-varying growth rates

# Appendix

## Stan Code

```{r, eval = FALSE}
data {
  int<lower = 0> N_obs;
  int country[N_obs];
  vector[N_obs] days;
  int new_cases[N_obs];
  int total_cases[N_obs];
  int<lower = 0> N_countries;
  vector[N_countries] pop;
}

parameters {
  vector<lower = 0>[N_countries] beta;
  vector[N_countries] alpha;
  vector<lower = 0, upper = 1>[N_countries] maximum;
  
  real<lower = 0> mu_beta;
  real<lower = 0> sigma_sq_beta;
  
  real mu_alpha;
  real<lower = 0> sigma_sq_alpha;
  
  real<lower = 0> beta_a;
  real<lower = 0> beta_b;
  
}

transformed parameters {
  vector[N_obs] linear = alpha[country] + beta[country] .* days;
  # vector<lower = 0, upper = 1>[N_obs] rate;
  vector<lower = 0, upper = 1>[N_obs] difference;
  for (i in 1:N_obs) {
    difference[i] = beta[country[i]] * maximum[country[i]] * exp(-linear[i]) / square(exp(-linear[i]) + 1);
    # rate[i] = maximum[country[i]] / (1 + exp(-linear[i]));
  }
}

model {
  
  sigma_sq_beta ~ inv_chi_square(2);
  sigma_sq_alpha ~ inv_chi_square(2);
  
  maximum ~ beta(beta_a, beta_b);
  
  beta ~ normal(mu_beta, sigma_sq_beta);
  alpha ~ normal(mu_alpha, sigma_sq_alpha);
  
  
  new_cases ~ poisson(difference .* pop[country]);
  # total_cases ~ poisson(rate .* pop[country]);
}

```

